Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
worked on drought descriptions

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community. 
Candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth. 
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

*
In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs. 1, S2). We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [*Anderson-Teixeira et al., in review*; @twery_effects_1991].

Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).

The droughts differed in intensity and prior onset (Table S3, Fig. S2). 1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season. The 1977 drought was the least intense throughout the growing season. All droughts increased in intensity from May through August, but differed in the previous months. The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June. The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.*

>>>>

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth. Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community.

We identified the years with driest conditions during May-August, which stood out in the analysis of @helcoski_growing_2019 as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). {**ADD SPEI**} These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest. 
Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).

We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. $Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate increased growth. Pointer years were identified using the pointRes package [@R-pointRes] in R. Four years met our criteria: 1966, 1977, 1991, and 1999. We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period. Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [*Anderson-Teixeira et al., in review*; @twery_effects_1991].

Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and prior onset (Fig. S2, Table S3). 1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season. The 1977 drought was the least intense throughout the growing season. All droughts increased in intensity from May through August, but differed in the previous months. The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June. The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
